Simple repression draft

### 2020-06-05 doc/section_03_beyond_means.tex
Finish appendix on mean and Fanos for nonbursty models
Not quite final pass on everything, Appdx S3 still needs more

Before we can tackle simple repression, we need an adequate phenomenological
model of constitutive expression. The literature abounds with options from which
we can choose, and we show several potential kinetic models for constitutive
promoters in~\fig{fig2:constit_cartoons}. Let us consider the suitability of
each model for our purposes in turn.

@@

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons} and assumes that transcripts are
produced as a Poisson process from a single promoter state. This is the picture
from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
study of gene expression noise over a series of promoters designed to have
different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescent spots, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For disproving the ``universal noise curve'' from So et
al~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
Since $\gamma$ is independently known from other experiments, and the mean mRNA
copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
model is not excessively complex for the data at hand. But for many interesting
questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
means and variances alone is insufficient, and a description of the full
distribution of molecular counts is necessary. For this we need a (slightly)
more complex model than model 1 that would allow us to incorporate the
non-Poissonian features of constitutive promoters directly into a master
equation formulation.

@@

Our second model of constitutive transcription posits an architecture in which
the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons},
transcription initiation results in a state transition from the bound to the
unbound state, reflecting the microscopic reality that an RNAP that has begun to
elongate a transcript is no longer available at the start site to begin another.
As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\end{align}
The problem with this picture is that the Fano factor is always $<1$. To make
 contact with the experimental reality of $\nu>1$ as shown in
 Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
 While this model adds an appealing element of microscopic reality, we are
 forced to reject it as the additional complexity is unable to capture the
 phenomenology of interest. Obviously the promoter state does in fact proceed
 through cycles of RNAP binding, initiating, and elongating, but it seems that
 the super-Poissonian noise in mRNA copy number we want to model must be
 governed by other features of the system.

@@

How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
that once RNAP initially binds the promoter region, a multitude of distinct
steps occur sequentially before RNAP finally escapes into the elongation phase.
Perhaps adding some of this mechanistic detail as shown in model 3 of
Figure~\ref{fig2:constit_cartoons} might rescue the previous model. The next
simplest refinement of that model could consider open complex formation and
promoter escape as separate steps rather than as a single effective step. In
other words, we construct model 3 by adding a single extra state to model 2, and
we will label the two RNAP-bound states as the closed and open complexes,
despite the true biochemical details certainly being more complex. For example,
earlier work extended this model by adding an additional repressor bound state
and did not explicitly consider the limit with no repressor that we analyze
here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
all the relevant biochemical detail; this is an exploratory search for the
important features that a model needs to include to square with the known
experimental reality of constitutive expression.

Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
has Fano factor $\nu<1$. This can be seen by starting with their Eq.~A1 for the
Fano factor of the analogous model with repressor and taking the limit as
repressor concentration goes to zero. After substantial algebra, the result is
\begin{align}
\nu = 1 - \frac{r k_O k_P^+}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + \gamma}
        {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + \gamma) + \gamma^2},
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

\marginpar{\it RP: very fun next paragraph - I like the argument, need to think more}
In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively should be more peaked with a smaller fractional width than an
exponential with the same mean.
\footnote{This can be made more precise with a result from~\cite{Stewart2007},
who showed that the convolution of multiple gamma distributions (of which the
exponential distribution is a special case) is, to a very good approximation,
also gamma distributed. Using their Eq.~2 for the distribution of the
convolution, with shape parameters set to 1 to give exponential distributions,
the total waiting time distribution has a ratio of variance to squared mean
$\sigma^2/\mu^2 = \sum_i k_i^2/\left(\sum_i k_i\right)^2$, where the $k_i$ are
the rates of the individual steps. Clearly this is less than 1 and therefore the
total waiting time distribution is narrower than the corresponding exponential.}
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
\footnote{This last step, while intuitive, can be argued by analogy to photon
statistics where antibunching gives rise to sub-Poissonian noise~\cite{Paul1982,
Zou1990}. Although loopholes exist, we do not expect they apply for our problem.
Nevertheless we refrain from elevating this cycle/sub-Poissonian equivalence to
a ``theorem.'' \mmnote{Antibunching is the ``obvious'' analogy for me but it'd
be totally out of left field for most readers, any suggestions on a more
comprehensible reference??}}
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.

@@

Inspired by~\cite{Razo-Mejia2020}, we revisit an analog of model 2, but as with
the models considered in \ref{section_02_means}, the interpretation of the two
states is changed. Rather than explicitly viewing them as RNAP bound and
unbound, we view them as ``active'' and ``inactive,'' which are able and unable
to initiate transcripts, respectively. We are noncommittal as to the microscopic
details of these states.

@@

Intuitively one might expect that, since transcripts are produced as a Poisson
process only when the promoter is in one of the two states in this model,
transcription initiations should now be ``bunched,'' in contrast to the
``anti-bunching'' of models 2 and 3 above. One might further guess that this
bunching would lead to super-Poissonian noise in the mRNA distribution over a
population of cells.  Indeed, as shown in Appendix~\ref{sec:non_bursty}, a
calculation of the Fano factor produces
\begin{align}
\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\end{align}
which is strictly greater than 1, verifying the above intuition. Note we have
dropped the $P$ label on the promoter switching rates to emphasize that these
very likely do not represent kinetics of RNAP itself. This calculation can also
be sidestepped by noting that the model is mathematically equivalent to the
simple repression model from~\cite{Jones2014}, with states and rates relabeled
and reinterpreted.

@@

\marginpar{\it RP: best description I have ever seen of this.  Good job}
To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as reported in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016}\mmnote{should probably add some
more cites here}. Let us make this correspondence more precise. The mean dwell
time in the active state is $1/k^-$. While in this state, transcripts are
produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
produced before the system switches to the inactive state. Once in the inactive
state, the system dwells there for an average time $1/k^+$ before returning to
the active state and repeating the process. $r/k^-$ resembles an average burst
size, and $1/k^+$ resembles the time interval between burst events. More
precisely, $1/k^+$ is the mean time between the end of one burst and the start
of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
bursts is approximately exponentially distributed, with mean set by the only
timescale left in the problem, $1/k^+$.
\footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
$1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
the case for any of datasets from~\cite{Jones2014}.}

@@

In reflecting on these 5 models, the reader may feel that exploring a multitude
of potential models just to return to a very minimal phenomenological model of
bursty transcription may seem highly pedantic. But the purpose of the exercise
was to examine a host of models from the literature and understand why they are
insufficient, one way or another, for our purposes. Along the way we have
learned that the detailed kinetics of RNAP binding and initiating transcription
are probably irrelevant for setting the population distribution of mRNA. The
timescales are simply too fast, and as we will see later
in~\fig{fig:constit_post}, the noise seems to be governed by slower timescales.
Perhaps in hindsight this is not surprising: intuitively, the degradation rate
$\gamma$ sets the fundamental timescale for mRNA dynamics, and any other
processes that substantially modulate the mRNA distribution should not differ
from $\gamma$ by orders of magnitude.

>>>>

Before we can tackle simple repression, we need an adequate phenomenological
model of constitutive expression. The literature abounds with options from which
we can choose, and we show several potential kinetic models for constitutive
promoters in Figure~\ref{fig2:constit_cartoons}(A). Let us consider
the suitability of each model for our purposes in turn.

@@

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts are
produced as a Poisson process from a single promoter state. This is the picture
from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
study of gene expression noise over a series of promoters designed to have
different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescent spots, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For disproving the ``universal noise curve'' from So et.\
al.~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
Since $\gamma$ is independently known from other experiments, and the mean mRNA
copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
model is not excessively complex for the data at hand. But for many interesting
questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
means and variances alone is insufficient, and a description of the full
distribution of molecular counts is necessary. For this we need a (slightly)
more complex model than model 1 that would allow us to incorporate the
non-Poissonian features of constitutive promoters directly into a master
equation formulation.

@@

Our second model of constitutive transcription posits an architecture in which
the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
transcription initiation results in a state transition from the bound to the
unbound state, reflecting the microscopic reality that an RNAP that has begun to
elongate a transcript is no longer available at the start site to begin another.
As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\label{eq:model2_fano}
\end{align}
The problem with this picture is that the Fano factor is always $<1$. To make
 contact with the experimental reality of $\nu>1$ as shown in
 Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
 While this model adds an appealing element of microscopic reality, we are
 forced to reject it as the additional complexity is unable to capture the
 phenomenology of interest. Obviously the promoter state does in fact proceed
 through cycles of RNAP binding, initiating, and elongating, but it seems that
 the super-Poissonian noise in mRNA copy number we want to model must be
 governed by other features of the system.

@@

How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
that once RNAP initially binds the promoter region, a multitude of distinct
steps occur sequentially before RNAP finally escapes into the elongation phase.
Perhaps adding some of this mechanistic detail as shown in model 3 of
Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
simplest refinement of that model could consider open complex formation and
promoter escape as separate steps rather than as a single effective step. In
other words, we construct model 3 by adding a single extra state to model 2, and
we will label the two RNAP-bound states as the closed and open complexes,
despite the true biochemical details certainly being more complex. For example,
earlier work extended this model by adding an additional repressor bound state
and did not explicitly consider the limit with no repressor that we analyze
here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
all the relevant biochemical detail; this is an exploratory search for the
important features that a model needs to include to square with the known
experimental reality of constitutive expression.

Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
has Fano factor $\nu<1$. We again leave the algebraic details for
Appendix~\ref{sec:non_bursty} and merely state the result that
\begin{align}
\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + r + \gamma}
    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
\label{eq:model3_fano}
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively should be more peaked with a smaller fractional width than an
exponential with the same mean.
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
(In Appendix~\ref{sec:non_bursty} we present a more precise version
of the intuitive arguments in this paragraph.)
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.

@@

Inspired by~\cite{Razo-Mejia2020}, we next revisit an analog of model 2 in
Figure~\ref{fig2:constit_cartoons}(A), but as with
the analogous models considered in Section~\ref{section_02_means},
the interpretation of the two
states is changed. Rather than explicitly viewing them as RNAP bound and
unbound, we view them as ``active'' and ``inactive,'' which are able and unable
to initiate transcripts, respectively. We are noncommittal as to the microscopic
details of these states.

@@

Intuitively one might expect that, since transcripts are produced as a Poisson
process only when the promoter is in one of the two states in this model,
transcription initiations should now be ``bunched'' in time, in contrast to the
``anti-bunching'' of models 2 and 3 above. One might further guess that this
bunching would lead to super-Poissonian noise in the mRNA distribution over a
population of cells.  Indeed, as shown in Appendix~\ref{sec:non_bursty}, a
calculation of the Fano factor produces
\begin{align}
\nu &= 1 + \frac{r k^-}{(k^+ + k^- + \gamma)(k^+ + k^-)},
\label{eq:model4_fano}
\end{align}
which is strictly greater than 1, verifying the above intuition. Note we have
dropped the $P$ label on the promoter switching rates to emphasize that these
very likely do not represent kinetics of RNAP itself. This calculation can also
be sidestepped by noting that the model is mathematically equivalent to the
simple repression model from~\cite{Jones2014}, with states and rates relabeled
and reinterpreted.

@@

% \marginpar{\it RP: best description I have ever seen of this.  Good job}
To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as reported in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016}.
%\mmnote{should probably add some more cites here}.
Let us make this correspondence more precise. The mean dwell
time in the active state is $1/k^-$. While in this state, transcripts are
produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
produced before the system switches to the inactive state. Once in the inactive
state, the system dwells there for an average time $1/k^+$ before returning to
the active state and repeating the process. $r/k^-$ resembles an average burst
size, and $1/k^+$ resembles the time interval between burst events. More
precisely, $1/k^+$ is the mean time between the end of one burst and the start
of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
bursts is approximately exponentially distributed, with mean set by the only
timescale left in the problem, $1/k^+$.
% \footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
% $1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
% the case for any of datasets from~\cite{Jones2014}.}

@@

In reflecting on these 5 models, the reader may feel that exploring a multitude
of potential models just to return to a very minimal phenomenological model of
bursty transcription may seem highly pedantic. But the purpose of the exercise
was to examine a host of models from the literature and understand why they are
insufficient, one way or another, for our purposes. Along the way we have
learned that the detailed kinetics of RNAP binding and initiating transcription
are probably irrelevant for setting the population distribution of mRNA. The
timescales are simply too fast, and as we will see later in
Figures~\ref{fig:constit_post_full} and~\ref{fig4:repressed_post_full},
the noise seems to be governed by slower timescales.
Perhaps in hindsight this is not surprising: intuitively, the degradation rate
$\gamma$ sets the fundamental timescale for mRNA dynamics, and any other
processes that substantially modulate the mRNA distribution should not differ
from $\gamma$ by orders of magnitude.
